---
title: "The Protein Leak"
subtitle: "Mapping the Fragility of West African Food Security Through OSINT, AIS Tracking, and Bayesian Inference"
author: "Kristian Vepsäläinen"
date: today
categories: [OSINT, Bayesian, West Africa, Fisheries, Security, Causal Inference, English]
description: |
  Industrial trawlers ghost through West African EEZs with AIS transponders switched off.
  Fish stocks collapse. Gulf of Guinea piracy surges. Tens of thousands flee by boat toward
  Europe. This whitepaper traces the probabilistic causal chain — from stolen protein to
  regional destabilization — using four years of real Global Fishing Watch AIS data (2022–2025),
  Bayesian changepoint detection, and distributional causal inference.
  The world is a distribution. So is the risk.
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    code-fold: true
    code-tools: true
    fig-cap-location: bottom
    fig-width: 10
    fig-height: 6
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---
::: {.callout-note appearance="simple"}
**Download:** [PDF version](protein-leak.pdf) · [Source on GitHub](https://github.com/KristianVepsalainen/KristianVepsalainen.github.io/blob/main/whitepapers/protein-leak.qmd) 
:::
```{r setup}
#| label: setup
#| include: false

library(tidyverse)
library(wbstats)
library(bcp)
library(rstanarm)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(patchwork)
library(scales)
library(glue)
library(here)
# install: install.packages("gfwr",
#   repos = c("https://globalfishingwatch.r-universe.dev","https://cran.r-project.org"))
library(gfwr)

set.seed(42)


DATA_DIR     <- here("data", "protein-leak")
dir.create(DATA_DIR, showWarnings = FALSE, recursive = TRUE)

COL_RED    <- "#e63946"
COL_TEAL   <- "#2a9d8f"
COL_ORANGE <- "#f4a261"
COL_NAVY   <- "#1d3557"
COL_BLUE   <- "#457b9d"
COL_BG     <- "#0d1117"
COL_LIGHT  <- "#e8e8e8"
COL_MID    <- "#8888aa"

theme_protein <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.background  = element_rect(fill = COL_BG,    color = NA),
      panel.background = element_rect(fill = "#111827", color = NA),
      panel.grid.major = element_line(color = "#1e2d3d", linewidth = 0.4),
      panel.grid.minor = element_line(color = "#151f2a", linewidth = 0.2),
      text             = element_text(color = COL_LIGHT, family = "sans"),
      axis.text        = element_text(color = COL_MID),
      plot.title       = element_text(color = COL_LIGHT, face = "bold",
                                      size  = base_size * 1.3),
      plot.subtitle    = element_text(color = COL_MID,   size  = base_size * 0.95),
      plot.caption     = element_text(color = COL_MID,   size  = base_size * 0.75,
                                      hjust = 0),
      legend.background = element_rect(fill = COL_BG,   color = NA),
      legend.text       = element_text(color = COL_LIGHT),
      legend.title      = element_text(color = COL_LIGHT),
      strip.text        = element_text(color = COL_LIGHT, face = "bold"),
      strip.background  = element_rect(fill = "#182033", color = NA)
    )
}
theme_set(theme_protein())

GFW_YEARS <- 2022:2025

west_af_sf <- ne_countries(continent = "Africa", scale = "medium",
                           returnclass = "sf") |>
  filter(name %in% c("Mauritania", "Senegal", "Gambia", "Guinea-Bissau",
                      "Guinea", "Sierra Leone", "Liberia", "Côte d'Ivoire",
                      "Ghana", "Togo", "Benin", "Nigeria", "Cameroon"))
```

---

::: {.callout-note}
## Executive Summary

Every year, industrial fishing fleets extract an estimated **1.5–2.6 million tonnes**
of fish from West African waters that never appear in any official landing record.

This whitepaper presents three interlocking findings, built on four years of real
Global Fishing Watch AIS data (2022–2025):

1. **Dark vessels are trackable.** High-confidence AIS disabling events cluster inside
   West African EEZs. A Kalman filter applied to last known position and implied speed
   constrains the search area — even without initial heading data.

2. **The depletion is structural, not cyclical.** Bayesian changepoint analysis places
   the inflection in West African catches in 2007–2010 with 94% posterior probability.

3. **The causal chain runs to piracy and migration.** A Bayesian negative-binomial
   regression shows each 100,000-tonne decline in annual EEZ catch is associated with
   22–41 additional Gulf of Guinea piracy incidents per year (90% credible interval).

**The world is a distribution. So is the risk.**
:::

::: {.callout-important}
## Data requirements

Section 2 uses the **Global Fishing Watch API** via the `gfwr` R package (v3.0).
Register for a free token at [globalfishingwatch.org/our-apis/tokens](https://globalfishingwatch.org/our-apis/tokens)
and add `GFW_TOKEN="your_token_here"` to `.Renviron`. In GitHub Actions, store as
secret `GFW_TOKEN`. All API responses are cached as `.rds` files and committed to
the repository — CI never calls the API directly.
:::

---

## The Problem: When Protein Vanishes, Violence Fills the Vacuum

```{r wb-malnutrition}
#| label: wb-malnutrition
#| cache: true

wa_iso3 <- c("MRT","SEN","GMB","GNB","GIN","SLE",
             "LBR","CIV","GHA","TGO","BEN","NGA","CMR")

wb_path <- file.path(DATA_DIR, "wb_malnutrition.rds")
if (!file.exists(wb_path)) {
  wb_raw <- wb_data(
    country     = wa_iso3,
    indicator   = c(undernourishment = "SN.ITK.DEFC.ZS",
                    gdp_pc           = "NY.GDP.PCAP.CD",
                    population       = "SP.POP.TOTL"),
    start_date  = 2000, end_date = 2023, return_wide = TRUE
  )
  saveRDS(wb_raw, wb_path)
} else { wb_raw <- readRDS(wb_path) }

stopifnot(is.data.frame(wb_raw), nrow(wb_raw) > 0,
          "undernourishment" %in% names(wb_raw))

wb_clean <- wb_raw |>
  filter(!is.na(undernourishment)) |>
  rename(year = date) |>
  mutate(year = as.integer(year))
```

Fish is not a luxury in West Africa. For coastal communities from Mauritania to
Nigeria, it provides **50–80% of daily animal protein**. Yet every morning, hundreds
of industrial vessels trawl the same waters — many without authorisation, and many
more misreporting their catch.

```{r plot-undernourishment}
#| label: plot-undernourishment
#| fig-cap: "Prevalence of undernourishment across West African coastal nations, 2000–2023. The ribbon shows the interquartile range; dashed line = regional median. Source: World Bank WDI."

annual_summary <- wb_clean |>
  group_by(year) |>
  summarise(median_pct = median(undernourishment, na.rm = TRUE),
            q25 = quantile(undernourishment, 0.25, na.rm = TRUE),
            q75 = quantile(undernourishment, 0.75, na.rm = TRUE),
            .groups = "drop")

ggplot() +
  geom_ribbon(data = annual_summary,
              aes(x = year, ymin = q25, ymax = q75),
              fill = COL_TEAL, alpha = 0.2) +
  geom_line(data = wb_clean |> filter(!iso3c %in% c("SLE","GIN","NGA","SEN")),
            aes(x = year, y = undernourishment, group = iso3c),
            colour = COL_MID, alpha = 0.35, linewidth = 0.5) +
  geom_line(data = wb_clean |> filter(iso3c %in% c("SLE","GIN","NGA","SEN")),
            aes(x = year, y = undernourishment, colour = country, group = iso3c),
            linewidth = 0.9) +
  geom_line(data = annual_summary, aes(x = year, y = median_pct),
            colour = COL_TEAL, linewidth = 1.2, linetype = "dashed") +
  scale_colour_manual(
    values = c("Sierra Leone" = COL_RED, "Guinea" = COL_ORANGE,
               "Nigeria" = COL_BLUE, "Senegal" = "#a8dadc"),
    name = NULL) +
  scale_x_continuous(breaks = seq(2000, 2023, 5)) +
  scale_y_continuous(labels = label_percent(scale = 1), limits = c(0, NA)) +
  labs(title   = "Undernourishment in West African Coastal Nations",
       subtitle = "IQR ribbon; dashed = regional median",
       x = NULL, y = "% of population undernourished",
       caption = "Source: World Bank WDI (SN.ITK.DEFC.ZS).")
```

### The Distribution of the Leak

```{r iuu-distribution}
#| label: iuu-distribution
#| fig-cap: "Distribution of annual IUU fishing volume. The ribbon is not measurement noise — it is epistemic uncertainty about what is, by definition, unreported."

alpha_iuu <- 4.8; beta_iuu <- 8.2
stopifnot(abs(alpha_iuu / (alpha_iuu + beta_iuu) - 0.37) < 0.01)
total_catch_reconstructed <- 3.8e6
iuu_samples <- rbeta(50000, alpha_iuu, beta_iuu) * total_catch_reconstructed
q_iuu <- quantile(iuu_samples, c(0.05, 0.25, 0.50, 0.75, 0.95))

ggplot(tibble(iuu_mt = iuu_samples), aes(x = iuu_mt / 1e6)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 80, fill = COL_TEAL, colour = NA, alpha = 0.7) +
  geom_vline(xintercept = q_iuu["50%"] / 1e6, colour = COL_ORANGE, linewidth = 1.0) +
  geom_vline(xintercept = c(q_iuu["5%"], q_iuu["95%"]) / 1e6,
             colour = COL_RED, linewidth = 0.7, linetype = "dashed") +
  annotate("text", x = q_iuu["50%"]/1e6 + 0.03, y = Inf, vjust = 1.5,
           label = glue("Median\n{round(q_iuu['50%']/1e6,2)}M t"),
           colour = COL_ORANGE, size = 3.5, hjust = 0) +
  annotate("text", x = q_iuu["5%"]/1e6 - 0.03, y = Inf, vjust = 1.5,
           label = glue("5th pct\n{round(q_iuu['5%']/1e6,2)}M t"),
           colour = COL_RED, size = 3.0, hjust = 1) +
  annotate("text", x = q_iuu["95%"]/1e6 + 0.03, y = Inf, vjust = 1.5,
           label = glue("95th pct\n{round(q_iuu['95%']/1e6,2)}M t"),
           colour = COL_RED, size = 3.0, hjust = 0) +
  scale_x_continuous(labels = label_number(suffix = "M t", accuracy = 0.1)) +
  labs(title   = "Distribution of Annual IUU Catch, West African EEZs",
       subtitle = "Beta(4.8, 8.2) × 3.8M t total reconstructed catch",
       x = "Annual IUU volume (million metric tonnes)", y = "Probability density",
       caption = "Prior: Sea Around Us (Pauly et al.); MRAG/Poseidon (2014); Belhabib et al. (2015)")
```

---

## Dark Vessels: Four Years of Real AIS Data

### Setup and Data Retrieval

```{r gfw-setup}
#| label: gfw-setup

gfw_token <- Sys.getenv("GFW_TOKEN")
if (nchar(gfw_token) < 10) {
  stop(paste(
    "GFW_TOKEN not found. Register at https://globalfishingwatch.org/our-apis/tokens",
    "and add GFW_TOKEN to .Renviron. CI uses cached .rds files."
  ))
}

wa_sf <- sf::st_bbox(
  c(xmin = -22, ymin = -2, xmax = 10, ymax = 22), crs = 4326
) |> sf::st_as_sfc() |> sf::st_as_sf()

stopifnot(inherits(wa_sf, "sf"))
```

```{r gfw-effort}
#| label: gfw-effort
#| cache: true

# Fetch fishing effort year-by-year; cache each year separately
effort_raw <- map_dfr(GFW_YEARS, function(yr) {
  path <- file.path(DATA_DIR, glue("gfw_effort_wa_{yr}.rds"))
  if (!file.exists(path)) {
    message(glue("Fetching fishing effort {yr}..."))
    raw <- gfw_ais_fishing_hours(
      spatial_resolution  = "LOW",
      temporal_resolution = "YEARLY",
      group_by            = "FLAG",
      start_date          = glue("{yr}-01-01"),
      end_date            = glue("{yr + 1}-01-01"),
      region              = wa_sf,
      region_source       = "USER_SHAPEFILE"
    ) |> mutate(data_year = yr)
    saveRDS(raw, path)
    raw
  } else {
    readRDS(path)
  }
})

stopifnot(is.data.frame(effort_raw), nrow(effort_raw) > 0)

effort_clean <- effort_raw |>
  rename_with(tolower) |>
  rename_with(~ str_replace_all(., " ", "_")) |>
  rename(
    fishing_hours = apparent_fishing_hours,
    year          = data_year
  ) |>
  filter(!is.na(fishing_hours), fishing_hours > 0)

stopifnot(all(c("lat", "lon", "fishing_hours", "year") %in% names(effort_clean)))
```

```{r gfw-gap-events}
#| label: gfw-gap-events
#| cache: true

gaps_raw <- map_dfr(GFW_YEARS, function(yr) {
  path <- file.path(DATA_DIR, glue("gfw_gaps_wa_{yr}.rds"))
  if (!file.exists(path)) {
    message(glue("Fetching GAP events {yr}..."))
    raw <- gfw_event(
      event_type    = "GAP",
      start_date    = glue("{yr}-01-01"),
      end_date      = glue("{yr + 1}-01-01"),
      vessel_types  = "FISHING",
      duration      = 6 * 60,
      region        = wa_sf,
      region_source = "USER_SHAPEFILE",
      quiet         = FALSE
    )
    saveRDS(raw, path)
    raw
  } else {
    readRDS(path)
  }
})

stopifnot(is.data.frame(gaps_raw), nrow(gaps_raw) > 0)
```

```{r parse-gaps}
#| label: parse-gaps

gaps <- gaps_raw |>
  rename(start_lat = lat, start_lon = lon) |>   # <-- tämä puuttui
  mutate(
    start      = as.POSIXct(start, tz = "UTC"),
    end        = as.POSIXct(end,   tz = "UTC"),
    year       = as.integer(format(start, "%Y")),
    duration_h = as.numeric(difftime(end, start, units = "hours")),
    # NULL-turvallinen unnestaus (reappearance saattaa puuttua avoimista gapeista)
    end_lat = map_dbl(event_info, ~ {
      v <- .x$onPosition$lat
      if (is.null(v)) NA_real_ else as.numeric(v)
    }),
    end_lon = map_dbl(event_info, ~ {
      v <- .x$onPosition$lon
      if (is.null(v)) NA_real_ else as.numeric(v)
    }),
    implied_speed_kn = map_dbl(event_info, ~ {
      v <- .x$impliedSpeedKnots
      if (is.null(v)) NA_real_ else as.numeric(v)
    }),
    distance_km = map_dbl(event_info, ~ {
      v <- .x$distanceKm
      if (is.null(v)) NA_real_ else as.numeric(v)
    })
  ) |>
  filter(duration_h >= 6, duration_h <= 240,
         year %in% GFW_YEARS)
stopifnot(all(gaps$duration_h >= 6))

n_gaps     <- nrow(gaps)
n_flags    <- n_distinct(gaps$vessel_flag, na.rm = TRUE)
median_dur <- median(gaps$duration_h, na.rm = TRUE)
```

### Where the Fleets Fish — and Where They Disappear

```{r plot-effort-facet}
#| label: plot-effort-facet
#| fig-cap: "Annual fishing effort (apparent fishing hours, 2022–2025) in West African waters. Each panel is one year; each cell is 0.1° × 0.1°. Comparing panels reveals whether fishing pressure is shifting geographically or intensifying. Source: Global Fishing Watch."
#| fig-height: 7

effort_grid <- effort_clean |>
  group_by(year, lat, lon) |>
  summarise(fishing_hours = sum(fishing_hours, na.rm = TRUE), .groups = "drop")

# Shared colour scale across all years
max_h <- quantile(effort_grid$fishing_hours, 0.99)

ggplot() +
  geom_tile(data = effort_grid,
            aes(x = lon, y = lat,
                fill = pmin(log1p(fishing_hours), log1p(max_h)))) +
  geom_sf(data = west_af_sf,
          fill = NA, colour = COL_LIGHT, linewidth = 0.25, alpha = 0.7) +
  scale_fill_gradient(
    low = COL_NAVY, high = COL_ORANGE,
    name = "log(hours + 1)",
    na.value = "transparent"
  ) +
  coord_sf(xlim = c(-22, 10), ylim = c(-2, 22)) +
  facet_wrap(~year, nrow = 2) +
  labs(title   = "Fishing Effort in West African Waters, 2022–2025",
       subtitle = "Colour scale identical across all panels; orange = high effort",
       x = "Longitude", y = "Latitude",
       caption = "Source: GFW apparent fishing effort (AIS-based).")
```

```{r plot-effort-top-flags}
#| label: plot-effort-top-flags
#| fig-cap: "Fishing effort by top flag states, 2022–2025. Lines reveal whether individual flag states are expanding or contracting their presence in West African waters — a proxy for IUU risk trends."

top_flags_overall <- effort_clean |>
  group_by(flag) |>
  summarise(total_h = sum(fishing_hours, na.rm = TRUE), .groups = "drop") |>
  slice_max(total_h, n = 7) |>
  pull(flag)

effort_ts <- effort_clean |>
  filter(flag %in% top_flags_overall) |>
  group_by(year, flag) |>
  summarise(fishing_hours = sum(fishing_hours, na.rm = TRUE), .groups = "drop")

flag_palette <- setNames(
  c(COL_RED, COL_ORANGE, COL_TEAL, COL_BLUE, "#a8dadc", "#e9c46a", COL_MID),
  top_flags_overall
)

ggplot(effort_ts, aes(x = year, y = fishing_hours / 1000,
                       colour = flag, group = flag)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_colour_manual(values = flag_palette, name = "Flag state") +
  scale_x_continuous(breaks = GFW_YEARS) +
  scale_y_continuous(labels = label_number(suffix = "k h")) +
  labs(title   = "Fishing Effort Trends by Flag State, 2022–2025",
       subtitle = "Top 7 flag states by total effort in West African waters",
       x = NULL, y = "Apparent fishing hours (thousands)",
       caption = "Source: GFW AIS fishing effort via gfwr.")
```

### The Distribution of Disappearances

```{r plot-gap-duration-violin}
#| label: plot-gap-duration-violin
#| fig-cap: "Distribution of AIS disabling event durations by year, fishing vessels in West African waters (2022–2025). Wider violins = more events; longer tails = longer dark periods. A rightward shift in the tail over years would indicate that IUU vessels are becoming bolder."

ggplot(gaps |> filter(duration_h <= 120),
       aes(x = factor(year), y = duration_h, fill = factor(year))) +
  geom_violin(alpha = 0.7, colour = NA, scale = "width") +
  geom_boxplot(width = 0.12, colour = COL_LIGHT,
               outlier.colour = COL_RED, outlier.size = 0.8,
               fill = COL_BG, alpha = 0.8) +
  scale_fill_manual(
    values = c("2022" = COL_NAVY, "2023" = COL_BLUE,
               "2024" = COL_TEAL, "2025" = COL_ORANGE),
    guide  = "none"
  ) +
  scale_y_continuous(breaks = c(6, 12, 24, 48, 72, 96, 120)) +
  labs(title   = "Dark Period Duration by Year",
       subtitle = glue("{scales::comma(n_gaps)} total high-confidence AIS disabling events"),
       x = NULL, y = "Gap duration (hours)",
       caption = "Source: GFW Events API (gfwr 3.0). Fishing vessels; min. 6h; truncated at 120h for display.")
```

```{r plot-gap-flag-timeseries}
#| label: plot-gap-flag-timeseries
#| fig-cap: "AIS disabling events by flag state, 2022–2025. A rising line indicates a flag state's vessels are going dark more frequently in West African EEZs — a direct indicator of IUU risk. A falling line may indicate genuine compliance improvement or shift to other evasion methods."

top_gap_flags <- gaps |>
  filter(!is.na(vessel_flag)) |>
  count(vessel_flag, name = "n_total") |>
  slice_max(n_total, n = 7) |>
  pull(vessel_flag)

gaps_flag_ts <- gaps |>
  filter(!is.na(vessel_flag), vessel_flag %in% top_gap_flags) |>
  count(year, vessel_flag)

gap_flag_palette <- setNames(
  c(COL_RED, COL_ORANGE, COL_TEAL, COL_BLUE, "#a8dadc", "#e9c46a", COL_MID),
  top_gap_flags
)

ggplot(gaps_flag_ts, aes(x = year, y = n, colour = vessel_flag, group = vessel_flag)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3.5) +
  geom_text(data = gaps_flag_ts |> filter(year == max(GFW_YEARS)),
            aes(label = vessel_flag),
            hjust = -0.15, size = 3.2, fontface = "bold") +
  scale_colour_manual(values = gap_flag_palette, guide = "none") +
  scale_x_continuous(breaks = GFW_YEARS,
                     limits = c(min(GFW_YEARS), max(GFW_YEARS) + 0.6)) +
  labs(title   = "AIS Disabling Events by Flag State, 2022–2025",
       subtitle = "Top 7 flag states; high-confidence intentional disabling; fishing vessels only",
       x = NULL, y = "Annual dark events (count)",
       caption = "Source: GFW Events API. Each event meets GFW's ≥14 satellite-position pre-gap criterion.")
```

```{r plot-gap-map-facet}
#| label: plot-gap-map-facet
#| fig-cap: "Spatial distribution of AIS disabling event start positions by year (2022–2025). Long gaps (≥24h, red) indicate deeper EEZ penetration. Persistent hot spots across years mark the most exploited fishing grounds."
#| fig-height: 8

gaps_map <- gaps |>
  filter(!is.na(start_lon), !is.na(start_lat)) |>
  mutate(duration_cat = if_else(duration_h >= 24, "≥24h (deep)", "<24h (short)"))

ggplot() +
  geom_sf(data = west_af_sf,
          fill = "#182033", colour = "#2a3a50", linewidth = 0.35) +
  geom_point(
    data   = gaps_map |> filter(duration_cat == "<24h (short)"),
    aes(x  = start_lon, y = start_lat, size = duration_h),
    colour = COL_TEAL, alpha = 0.3, shape = 16
  ) +
  geom_point(
    data   = gaps_map |> filter(duration_cat == "≥24h (deep)"),
    aes(x  = start_lon, y = start_lat, size = duration_h),
    colour = COL_RED, alpha = 0.5, shape = 16
  ) +
  scale_size_continuous(range = c(0.5, 4), name = "Duration (h)", guide = "none") +
  coord_sf(xlim = c(-22, 10), ylim = c(-2, 22)) +
  facet_wrap(~year, nrow = 2) +
  annotate("point", x = -21, y = -1, colour = COL_TEAL, size = 2) +
  annotate("text",  x = -20.3, y = -1, label = "< 24h",
           colour = COL_TEAL, size = 2.8, hjust = 0) +
  annotate("point", x = -21, y = -1.8, colour = COL_RED, size = 2) +
  annotate("text",  x = -20.3, y = -1.8, label = "≥ 24h",
           colour = COL_RED, size = 2.8, hjust = 0) +
  labs(title   = "Where Fishing Vessels Go Dark, 2022–2025",
       subtitle = "Each point = one AIS disabling event start position (high-confidence, ≥6h)",
       x = "Longitude", y = "Latitude",
       caption = "Source: GFW Events API. Comparing panels reveals whether hot spots persist, migrate, or intensify.")
```

### Dead Reckoning with a Kalman Filter

When a vessel goes dark, we know its last position and — from the gap event — the
average speed over the missing period. We do **not** have the initial heading from
the GFW API. This matters for the model: without heading, the uncertainty envelope
is a growing circle rather than a directed ellipse.

This is the **honest** result. If initial course data were available (e.g., from
satellite AIS before the gap), the search area would shrink dramatically. We state
this explicitly as a limitation and policy recommendation.

```{r kalman-implementation}
#| label: kalman-implementation

run_kalman_filter <- function(
    lon_obs, lat_obs,
    dt         = 1,
    sigma_proc = 0.04,
    sigma_obs  = 0.001
) {
  stopifnot(length(lon_obs) == length(lat_obs))
  stopifnot(dt > 0, sigma_proc > 0, sigma_obs > 0)

  n     <- length(lon_obs)
  F_mat <- matrix(c(1,0,dt,0, 0,1,0,dt, 0,0,1,0, 0,0,0,1),
                  nrow = 4, byrow = TRUE)
  H_mat <- matrix(c(1,0,0,0, 0,1,0,0), nrow = 2, byrow = TRUE)
  Q_mat <- sigma_proc^2 * diag(c(dt^2, dt^2, 1, 1))
  R_mat <- sigma_obs^2  * diag(2)

  state_mean <- matrix(NA_real_, n, 4,
                       dimnames = list(NULL, c("lon","lat","vlon","vlat")))
  state_sd   <- matrix(NA_real_, n, 2,
                       dimnames = list(NULL, c("sd_lon","sd_lat")))

  idx0 <- which(!is.na(lon_obs))[1]
  stopifnot(!is.na(idx0))
  x <- c(lon_obs[idx0], lat_obs[idx0], 0, 0)
  P <- diag(c(0.01, 0.01, 0.1, 0.1))

  for (i in seq_len(n)) {
    x <- as.vector(F_mat %*% x)
    P <- F_mat %*% P %*% t(F_mat) + Q_mat
    if (!is.na(lon_obs[i]) && !is.na(lat_obs[i])) {
      z     <- c(lon_obs[i], lat_obs[i])
      S     <- H_mat %*% P %*% t(H_mat) + R_mat
      K     <- P %*% t(H_mat) %*% solve(S)
      x     <- x + as.vector(K %*% (z - H_mat %*% x))
      P     <- (diag(4) - K %*% H_mat) %*% P
    }
    state_mean[i, ] <- x
    state_sd[i, ]   <- c(sqrt(P[1,1]), sqrt(P[2,2]))
  }
  list(mean = as.data.frame(state_mean), sd = as.data.frame(state_sd))
}
```

```{r kalman-on-real-gaps}
#| label: kalman-on-real-gaps
#| cache: true

gaps_kf <- gaps |>
  filter(
    !is.na(start_lon), !is.na(start_lat),
    !is.na(end_lon),   !is.na(end_lat),
    !is.na(implied_speed_kn),
    duration_h >= 12,  duration_h <= 120
  )

gaps_kf <- slice_sample(gaps_kf, n = min(300, nrow(gaps_kf)))

stopifnot(nrow(gaps_kf) > 0)

# Process noise scaled per event by implied speed:
# sigma_proc = implied_speed_kn / 60  → after N hours, 2σ ≈ 2*(speed/60)*√N degrees
# This makes the uncertainty circle physically meaningful (proportional to vessel speed)

kf_results <- gaps_kf |>
  rowwise() |>
  mutate(
    n_steps    = ceiling(duration_h),
    sigma_proc = max(0.005, implied_speed_kn / 60),

    lon_obs = list(c(start_lon, rep(NA_real_, n_steps))),
    lat_obs = list(c(start_lat, rep(NA_real_, n_steps))),

    kf = list(run_kalman_filter(
      unlist(lon_obs), unlist(lat_obs),
      dt = 1, sigma_proc = sigma_proc, sigma_obs = 0.001
    )),

    pred_lon         = kf$mean$lon[n_steps + 1],
    pred_lat         = kf$mean$lat[n_steps + 1],
    pred_sd_lon      = kf$sd$sd_lon[n_steps + 1],
    pred_sd_lat      = kf$sd$sd_lat[n_steps + 1],

    # Error: distance from Kalman prediction to actual reappearance
    error_nm = sqrt(
      ((pred_lon - end_lon) * 60 * cos(start_lat * pi / 180))^2 +
      ((pred_lat - end_lat) * 60)^2
    ),
    # 2σ search radius (nm)
    uncert_2sigma_nm = sqrt(
      (pred_sd_lon * 60 * cos(start_lat * pi / 180))^2 +
      (pred_sd_lat * 60)^2
    ) * 2,

    within_95 = error_nm <= uncert_2sigma_nm
  ) |>
  ungroup() |>
  select(-lon_obs, -lat_obs, -kf)

stopifnot(nrow(kf_results) > 0)
stopifnot(all(c("error_nm","uncert_2sigma_nm","within_95") %in% names(kf_results)))

coverage_pct  <- round(mean(kf_results$within_95, na.rm = TRUE) * 100, 0)
median_err_nm <- round(median(kf_results$error_nm, na.rm = TRUE), 0)
```

```{r plot-kalman-validation}
#| label: plot-kalman-validation
#| fig-cap: "Kalman filter validation against real GFW gap events. Left: prediction error distribution. Right: how the 2σ search radius grows with gap duration — without initial heading, the uncertainty is a circle whose radius scales with implied speed × √time. This is the worst case; initial heading data would substantially narrow the ellipse."

coverage_pct  <- round(mean(kf_results$within_95, na.rm = TRUE) * 100, 0)
median_err_nm <- round(median(kf_results$error_nm, na.rm = TRUE), 0)

# Rajataan järjettömät outlierit pois (implied_speed-outlierit → valtava ympyrä)
kf_plot <- kf_results |>
  filter(error_nm < 500, uncert_2sigma_nm < 500)

p_err <- ggplot(kf_plot, aes(x = error_nm)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50, fill = COL_TEAL, colour = NA, alpha = 0.8) +
  geom_vline(xintercept = median_err_nm,
             colour = COL_ORANGE, linewidth = 1.1) +
  scale_x_continuous(labels = label_number(suffix = " nm"),
                     breaks = c(0, 50, 100, 200, 300, 400, 500)) +
  labs(
    title    = "Kalman Prediction Error",
    subtitle = glue("n = {nrow(kf_results)} • Median error {median_err_nm} nm \n • {coverage_pct}% within 2σ region"),
    x        = "Prediction error (nautical miles)",
    y        = "Density",
    caption  = NULL
  )

p_uncert <- ggplot(kf_plot, aes(x = duration_h, y = uncert_2sigma_nm)) +
  geom_point(aes(colour = within_95), alpha = 0.4, size = 1.2) +
  geom_smooth(method = "loess", colour = COL_ORANGE,
              fill = COL_ORANGE, alpha = 0.2, se = TRUE) +
  scale_colour_manual(
    values = c("TRUE" = COL_TEAL, "FALSE" = COL_RED),
    labels = c("TRUE" = "Within 2σ", "FALSE" = "Outside 2σ"),
    name   = NULL
  ) + 
  scale_x_continuous(breaks = c(12, 24, 48, 72, 96, 120)) +
  scale_y_continuous(labels = label_number(suffix = " nm"),
                     breaks = c(0, 100, 200, 300, 400, 500)) +
  labs(
    title    = "Search Radius Grows \n with Gap Duration",
    subtitle = "Without initial heading: \n circular; with heading: directed ellipse",
    x        = "Gap duration (hours)",
    y        = "2σ search radius (nm)",
    caption  = "Source: GFW Events API + Kalman filter.\n Outliers (>500 nm) excluded."
  )
p_err | p_uncert
```

The filter achieves **`r coverage_pct`% coverage** across `r nrow(kf_results)` real
gap events with a median prediction error of `r median_err_nm` nautical miles.
Because initial course is unavailable, this is a conservative (pessimistic) model —
the true search area, were course data available, would be substantially smaller.

---

## The Ecosystem Cascade: Fish Stocks as a Security Indicator

```{r fao-catch-data}
#| label: fao-catch-data

# FAO FishStat Area 34 catch estimates, West African EEZ nations (all flags)
# Source: FAO FishStat 2023 release + Sea Around Us IUU reconstruction
fao_catch <- tibble(
  year = 2000:2022,
  reported_kt = c(
    1452,1479,1518,1493,1381,
    1348,1277,1213,1151,1121,
    1088,1049,1021, 983, 951,
     921, 896, 869, 851, 839,
     821, 812, 796
  ),
  iuu_estimate_kt = c(
    290,310,340,320,290,
    350,390,420,450,460,
    480,500,520,540,560,
    580,590,600,610,620,
    625,630,635
  )
) |>
  mutate(total_kt = reported_kt + iuu_estimate_kt,
         iuu_share = iuu_estimate_kt / total_kt)

stopifnot(nrow(fao_catch) == 23, all(fao_catch$reported_kt > 0),
          all(fao_catch$iuu_share > 0 & fao_catch$iuu_share < 1))
```

```{r changepoint-analysis}
#| label: changepoint-analysis
#| cache: true

bcp_fit <- bcp(y = fao_catch$reported_kt, burnin = 1000, mcmc = 10000)
stopifnot(length(bcp_fit$posterior.prob) == nrow(fao_catch))

bcp_df <- tibble(
  year      = fao_catch$year,
  catch_kt  = fao_catch$reported_kt,
  post_prob = bcp_fit$posterior.prob,
  post_mean = bcp_fit$posterior.mean
)
peak_cp_year <- bcp_df$year[which.max(bcp_df$post_prob)]
peak_cp_prob <- round(max(bcp_df$post_prob) * 100, 0)
```

```{r plot-catch-changepoint}
#| label: plot-catch-changepoint
#| fig-cap: "West African EEZ reported catch (2000–2022) and Bayesian changepoint posterior probability. The structural break at 2007–2010 coincides with the peak expansion of Chinese distant-water fleets following the 2006 Forum on China-Africa Cooperation."

p_catch <- ggplot(bcp_df, aes(x = year)) +
  geom_area(aes(y = catch_kt / 1000), fill = COL_TEAL, alpha = 0.25) +
  geom_line(aes(y = catch_kt / 1000), colour = COL_TEAL, linewidth = 1.1) +
  geom_line(aes(y = post_mean / 1000),
            colour = COL_ORANGE, linewidth = 0.9, linetype = "dashed") +
  geom_vline(xintercept = peak_cp_year,
             colour = COL_RED, linewidth = 1.0, linetype = "dotted") +
  annotate("text", x = peak_cp_year + 0.3, y = 1.6,
           label  = glue("Changepoint\n{peak_cp_year} ({peak_cp_prob}% posterior)"),
           colour = COL_RED, hjust = 0, size = 3.5) +
  scale_x_continuous(breaks = seq(2000, 2022, 4)) +
  scale_y_continuous(labels = label_number(suffix = "M t", accuracy = 0.1)) +
  labs(title   = "West African EEZ Reported Catch: Structural Break",
       subtitle = "Teal = reported catch; dashed orange = posterior mean (bcp)",
       x = NULL, y = "Annual catch (million metric tonnes)",
       caption = "Source: FAO FishStat Area 34 (2023). bcp: Barry & Hartigan (1993).")

p_cp <- ggplot(bcp_df, aes(x = year, y = post_prob)) +
  geom_area(fill = COL_RED, alpha = 0.35) +
  geom_line(colour = COL_RED, linewidth = 1.0) +
  geom_vline(xintercept = peak_cp_year,
             colour = COL_RED, linewidth = 1.0, linetype = "dotted") +
  scale_x_continuous(breaks = seq(2000, 2022, 4)) +
  scale_y_continuous(labels = label_percent(), limits = c(0, 1)) +
  labs(x = "Year", y = "P(changepoint)")

p_catch / p_cp + plot_layout(heights = c(3, 1))
```

---

## The Domino Effect: From Empty Nets to Empty Boats to Empty Futures

```{r piracy-data}
#| label: piracy-data

# Gulf of Guinea piracy incidents
# Source: ICC International Maritime Bureau Annual Reports, 2005–2023
imb_piracy <- tibble(
  year      = 2005:2023,
  incidents = c(36,52,50,49,64,
                89,102,58,42,45,
                54,95,97,100,119,
                130,132,57,33)
)
stopifnot(nrow(imb_piracy) == 19, all(imb_piracy$incidents > 0))

combined_df <- fao_catch |>
  filter(year >= 2005) |>
  inner_join(imb_piracy, by = "year") |>
  mutate(catch_scaled = scale(reported_kt)[, 1],
         year_scaled  = scale(year)[, 1])

stopifnot(nrow(combined_df) >= 15)
```

```{r bayesian-regression}
#| label: bayesian-regression
#| cache: true

model_piracy <- stan_glm(
  incidents ~ catch_scaled + year_scaled,
  data            = combined_df,
  family          = neg_binomial_2(),
  prior           = normal(0, 0.8, autoscale = FALSE),
  prior_intercept = normal(4.0, 1.0),
  prior_aux       = exponential(1),
  chains = 4, iter = 3000, warmup = 1000, seed = 42, refresh = 0
)

stopifnot(class(model_piracy)[1] == "stanreg")
rhats <- summary(model_piracy)[, "Rhat"]
stopifnot(all(rhats[is.finite(rhats)] < 1.05))
```

```{r plot-piracy-results}
#| label: plot-piracy-results
#| fig-cap: "Left: observed piracy incidents vs. 90% posterior predictive interval. Right: posterior of the fish catch coefficient — red area shows the proportion of draws below zero (fewer fish → more piracy)."

post_samples  <- as.data.frame(model_piracy)
beta_catch    <- post_samples$catch_scaled
prob_negative <- mean(beta_catch < 0)

fitted_vals <- posterior_predict(model_piracy, draws = 500)
fit_sum <- tibble(
  year       = combined_df$year,
  obs        = combined_df$incidents,
  fitted_med = apply(fitted_vals, 2, median),
  fitted_lo  = apply(fitted_vals, 2, quantile, 0.05),
  fitted_hi  = apply(fitted_vals, 2, quantile, 0.95)
)

expected_base <- exp(median(post_samples$`(Intercept)`))
delta_sd      <- -100000 / sd(combined_df$reported_kt)
irr_ci90      <- quantile(exp(delta_sd * beta_catch), c(0.05, 0.95))
add_lo <- round(abs((irr_ci90[1] - 1) * expected_base), 0)
add_hi <- round(abs((irr_ci90[2] - 1) * expected_base), 0)

p_fit <- ggplot(fit_sum, aes(x = year)) +
  geom_ribbon(aes(ymin = fitted_lo, ymax = fitted_hi),
              fill = COL_ORANGE, alpha = 0.25) +
  geom_line(aes(y = fitted_med), colour = COL_ORANGE, linewidth = 1.0) +
  geom_point(aes(y = obs), colour = COL_TEAL, size = 2.5) +
  labs(title   = "Gulf of Guinea Piracy:\n Observed vs Model",
       subtitle = "Orange = 90% posterior predictive interval",
       x = "Year", y = "Annual incidents (GoG)")

p_coef <- ggplot(tibble(beta = beta_catch), aes(x = beta)) +
  geom_area(data = ~ filter(., beta < 0),
            stat = "density", fill = COL_RED, alpha = 0.5) +
  geom_area(data = ~ filter(., beta >= 0),
            stat = "density", fill = COL_TEAL, alpha = 0.4) +
  geom_density(colour = COL_LIGHT, linewidth = 1.0) +
  geom_vline(xintercept = 0, colour = COL_ORANGE, linetype = "dashed") +
  annotate("text", x = median(beta_catch) - 0.05, y = Inf, vjust = 1.5,
           label = glue("{round(prob_negative*100,0)}% of\nposterior < 0"),
           colour = COL_RED, hjust = 1, size = 3.5) +
  labs(title   = "Posterior: Fish Catch → \n Piracy",
       subtitle = "β < 0 = fewer fish precedes more piracy",
       x = "Log-IRR coefficient", y = "Posterior density")

p_fit | p_coef
```

The posterior places **`r round(prob_negative * 100, 0)`% of probability on a negative
effect** of fish catch on piracy. A 100,000-tonne decline in annual catch is associated
with **`r add_lo`–`r add_hi` additional piracy incidents per year** (90% CI).

---

## Conclusions

1. **The IUU rate is a distribution**, not a number. The 90th percentile is
   substantially above the cited median.

2. **Dark vessels are constrained, not invisible.** Four years of real GFW data show
   that `r coverage_pct`% of actual reappearances fall within the conservative
   Kalman credible region. With initial course data, this envelope would narrow
   from a circle to a directed ellipse — a tractable search area for naval patrol.

3. **The fish-piracy connection holds** with `r round(prob_negative * 100, 0)`%
   posterior probability.

4. **The time trends (2022–2025) are the new story.** The GFW fishing effort and
   dark-event data reveal whether the pressure is intensifying, shifting coastline,
   or responding to enforcement. Point-in-time analysis misses all of this.

---

## Technical Appendix

**GFW Data.** All fishing effort and AIS gap events retrieved via `gfwr` v3.0
(GFW API v3). GAP events meet GFW's strict criteria: ≥14 satellite AIS positions
in the 12 hours prior to the gap.

**Kalman Filter.** Constant-velocity model; zero initial velocity (no course data
from API); process noise σ_proc = implied_speed_kn / 60 per event. This produces
a symmetric circular uncertainty envelope — conservative by design.

**Bayesian Changepoint.** Barry-Hartigan (1993) via `bcp`, 1,000 burnin + 10,000
MCMC iterations.

**Bayesian NB Regression.** `rstanarm::stan_glm()`, `neg_binomial_2()`, 4 chains
× 3,000 iterations, 1,000 warmup. All R-hat < 1.05.

---

## References

Barry, D., & Hartigan, J. A. (1993). A Bayesian analysis for change point problems.
*JASA*, 88(421), 309–319.

Belhabib, D., Sumaila, U. R., et al. (2015). Euros vs. Yuan.
*PLOS ONE*, 10(3), e0118351.

Global Fishing Watch. (2024). *gfwr: R package for accessing GFW APIs* (v3.0).

ICC International Maritime Bureau. (2005–2023). *Piracy and Armed Robbery Against
Ships: Annual Reports.*

MRAG & Poseidon. (2014). *Economic impacts of IUU fishing on developing countries.*
European Commission DG MARE.

Pauly, D., Zeller, D., et al. (2016). *Nature Communications*, 7, 10244.
