---
title: "One Hundredth of a Second"
subtitle: "A Bayesian forward simulation of aerodynamic drag at the 1980 Olympics"
author: "Kristian Vepsäläinen"
date: "2026-07"
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: c/t
    width: 1280
    height: 720
    transition: fade
    background-transition: fade
    incremental: false
    code-line-numbers: false
    highlight-style: github
    fig-align: center
    fig-cap-location: bottom
    title-slide-attributes:
      data-background-color: "#0b1d2e"
execute:
  echo: true
  warning: false
  message: false
  cache: true
knitr:
  opts_chunk:
    dev: "png"
    fig.width: 9
    fig.height: 5
    dpi: 150
---

```{r setup, include=FALSE}
library(tidyverse)
library(scales)

theme_talk <- function() {
  theme_minimal(base_size = 16, base_family = "sans") +
    theme(
      plot.title = element_text(face = "bold", size = 20),
      plot.subtitle = element_text(color = "grey30"),
      panel.grid.minor = element_blank(),
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA)
    )
}

ice_blue   <- "#3a7ca5"
ice_navy   <- "#0b1d2e"
ice_silver <- "#c9d6df"
accent     <- "#d9a441"
```

## {.center background-color="#0b1d2e"}

::: {style="color:#c9d6df; text-align:center;"}
<h1 style="color:#ffffff; font-size: 3.2em; line-height: 1.05; margin-bottom: 0.2em;">
One hundredth<br>of a second.
</h1>

<p style="font-size: 1.3em; color:#d9a441; margin-top: 0.5em;">
Lake Placid, 17 February 1980
</p>

<p style="margin-top: 2em; opacity: 0.85;">
Kristian Vepsäläinen &nbsp;·&nbsp; useR! 2026
</p>
:::

---

## The race {background-color="#0b1d2e" style="color:#c9d6df"}

::: {.columns}
::: {.column width="55%"}
**Men's 15 km cross-country skiing**

- Gold: **Thomas Wassberg** &nbsp; 41:57.63
- Silver: **Juha Mieto** &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; 41:57.64

<br>

::: {style="font-size: 2em; color:#d9a441; font-weight:700;"}
Δ = 0.01 s
:::

<br>

After 42 minutes of skiing, one hundredth of a second.
The IOC changed the rules. Times are now rounded to 0.1 s.
:::

::: {.column width="45%"}
::: {style="background:#13314a; padding: 1.5em; border-left: 3px solid #d9a441; font-size: 0.95em;"}
**The folk theory**

Juha Mieto had a beard.
A large, full beard.

For decades Finnish skiing folklore has asked:

*Did the beard cost him the gold?*
:::
:::
:::

---

## This is not a history talk

::: {.fragment .fade-in}
We **cannot** know what happened on 17 February 1980.
:::

::: {.fragment .fade-in}
<br>
We **can** ask a different, sharper question:
:::

::: {.fragment .fade-in}
<br>

> Under physically plausible assumptions,
> is **0.01 s** a *typical* outcome of additional aerodynamic drag,
> or an *extreme* one?

:::

::: {.fragment .fade-in}
<br>
This is a **modeling** question — and it is exactly the kind of question
R is good at.
:::

---

## Why this is a useR talk

This is a story about three things every applied modeller meets:

::: {.incremental}
1. **Sparse data.** We have three split times. That's it.
2. **Unknown nuisance parameters.** Course profile, wind, body shape, power output.
3. **A claim about a small margin.** "Negligible" vs. "plausible" — who decides?
:::

::: {.fragment}
The toolkit:

::: {style="font-size:1.1em;"}
**Bayesian forward simulation** + **R** + **honest priors** + **a single plot**.
:::
:::

---

## The plan

::: {.columns}
::: {.column width="33%"}
### 1. Physics

A drag-power model
of cross-country skiing.

Frontal area · drag coefficient · velocity³ · power.
:::

::: {.column width="33%"}
### 2. Priors

Honest, wide,
defensible ranges
for every unknown.

Not point estimates.
:::

::: {.column width="33%"}
### 3. Propagate

100,000 simulated races.

Look at the posterior
of the time loss
attributable to drag.
:::
:::

::: {.fragment style="margin-top:1.5em; text-align:center; font-size:1.1em; color:#3a7ca5;"}
No Stan. No JAGS. Just `runif`, `rnorm`, and a loop.
:::

---

## What we know

::: {style="font-size: 0.95em;"}

**Three split times for Juha Mieto** (sports-reference.com archive)

| Distance | Cumulative time | Split speed |
|---------:|----------------:|------------:|
|   5 km   |       13:25.00  |  6.21 m/s   |
|  10 km   |       25:51.97  |  6.70 m/s   |
|  15 km   |       41:57.64  |  5.18 m/s   |

<br>

**Venue:** Mt. Van Hoevenberg, Lake Placid, NY.
The trails still exist.

**Everything else is a prior.**
:::

---

## The physics

The power needed to overcome aerodynamic drag:

$$
P_\text{drag} \;=\; \tfrac{1}{2}\,\rho\,C_d\,A\,v^{3}
$$

::: {style="font-size: 0.95em; margin-top: 1em;"}
| Symbol | Meaning | Treatment |
|:-------|:--------|:----------|
| $\rho$ | Air density | Fixed at 1.2 kg/m³ |
| $C_d$ | Drag coefficient (body) | Prior: U(0.7, 0.9) |
| $A$ | Frontal area (body) | Fixed at 0.5 m² |
| $A_\text{beard}$ | Additional frontal area from beard | Prior: U(0.005, 0.015) m² |
| $v$ | Local velocity | Derived from splits + terrain |
| $P$ | Total power output | Prior: N(400, 30) W |
:::

::: {.fragment style="margin-top: 1em; color:#3a7ca5;"}
The beard adds frontal area. The body still has to push through air.
$\Delta P = \tfrac{1}{2}\rho C_d A_\text{beard} v^3 \cdot \alpha^{3}$ where $\alpha < 1$ accounts for slipstream shielding.
:::

---

## Anchoring v to the split times

The splits give us **average segment speeds**. We don't know how the speed
was distributed within each segment.

So we model it:

```{r anchor-demo, echo=TRUE, eval=FALSE}
profile     <- c(up = 0.4, flat = 0.3, down = 0.3)   # course mix
speed_mult  <- c(up = 0.7, flat = 1.0, down = 1.3)   # relative speeds

for (k in 1:3) {                                      # for each split
  v_obs <- v_segments[k]
  scale <- v_obs / sum(profile * speed_mult)          # match the split
  for (seg in names(profile)) {
    v <- rnorm(N, speed_mult[seg] * scale, 0.2)       # local v
    v[v < 1] <- 1
    v3_mean <- v3_mean + profile[seg] * v^3           # accumulate E[v³]
  }
}
```

::: {.fragment style="margin-top:0.5em; color:#3a7ca5;"}
Jensen's inequality matters: drag depends on $\mathbb{E}[v^3]$, not $\mathbb{E}[v]^3$.
Modelling within-segment variation is **not optional**.
:::

---

## The priors, visualized

```{r prior-viz, echo=FALSE, fig.height=4.2}
set.seed(2026)
N <- 100000
priors <- tibble(
  draw = 1:N,
  A_beard = runif(N, 0.005, 0.015),
  Cd_body = runif(N, 0.7, 0.9),
  P_total = rnorm(N, 400, 30),
  alpha   = runif(N, 0.25, 0.40)
) |> pivot_longer(-draw)

label_lookup <- c(
  A_beard = "Beard frontal area (m²)\nU(0.005, 0.015)",
  Cd_body = "Drag coefficient C_d\nU(0.7, 0.9)",
  P_total = "Total power P (W)\nN(400, 30)",
  alpha   = "Slipstream factor α\nU(0.25, 0.40)"
)
priors$name <- factor(priors$name, levels = names(label_lookup),
                      labels = label_lookup)

ggplot(priors, aes(value)) +
  geom_histogram(bins = 40, fill = ice_blue, color = "white", linewidth = 0.2) +
  facet_wrap(~name, scales = "free", nrow = 1) +
  labs(x = NULL, y = NULL) +
  theme_talk() +
  theme(axis.text.y = element_blank(),
        strip.text = element_text(face = "bold"))
```

::: {style="font-size: 0.9em; color:#555; margin-top:0.5em;"}
Wide on purpose. The goal is to ask: *does the conclusion survive
honest uncertainty?* — not to pretend we have a tight estimate.
:::

---

## The model — one screen of R

```{r model, echo=TRUE}
set.seed(1980)
N <- 100000;  rho <- 1.2;  A_body <- 0.5
T_total <- 41*60 + 57.64

v_segments <- c(5000 / (13*60 + 25.00),
                5000 / ((25*60 + 51.97) - (13*60 + 25.00)),
                5000 / ((41*60 + 57.64) - (25*60 + 51.97)))

profile    <- c(up = 0.4, flat = 0.3, down = 0.3)
speed_mult <- c(up = 0.7, flat = 1.0, down = 1.3)

A_beard <- runif(N, 0.005, 0.015)
Cd_body <- runif(N, 0.7, 0.9)
P_total <- rnorm(N, 400, 30)
alpha   <- runif(N, 0.25, 0.40)

v3_mean <- numeric(N)
for (k in 1:3) {
  v_obs <- v_segments[k]
  scale <- v_obs / sum(profile * speed_mult)
  for (seg in names(profile)) {
    v <- rnorm(N, speed_mult[seg] * scale, 0.2);  v[v < 1] <- 1
    v3_mean <- v3_mean + profile[seg] * v^3
  }
}
v3_mean <- v3_mean / 3

P_drag_body  <- 0.5 * rho * Cd_body * A_body * v3_mean
P_drag_beard <- P_drag_body * (A_beard / A_body) * alpha^3
delta_time   <- T_total * (P_drag_beard / P_total)

p_active        <- runif(N, 0.05, 0.20)   # fraction of race in beard-exposed posture
delta_time_real <- delta_time * p_active
```

---

## The result

```{r posterior-plot, echo=FALSE, fig.height=4.6}
df <- tibble(delta_time_real = delta_time_real)

p_loss <- mean(delta_time_real >= 0.01)

ggplot(df, aes(delta_time_real)) +
  geom_density(fill = ice_blue, color = ice_navy, alpha = 0.85, linewidth = 0.4) +
  geom_vline(xintercept = 0.01, linetype = "dashed",
             color = accent, linewidth = 1.1) +
  annotate("label", x = 0.01, y = Inf, vjust = 1.4,
           label = "0.01 s\nthe actual margin",
           color = accent, fill = "white", size = 4.5,
           label.size = 0, fontface = "bold") +
  scale_x_continuous(labels = label_number(accuracy = 0.01, suffix = " s"),
                     limits = c(0, quantile(df$delta_time_real, 0.995))) +
  labs(
    title = "Posterior of time loss attributable to beard drag",
    subtitle = sprintf("100,000 simulated races · P(Δt ≥ 0.01 s) = %.2f",
                       p_loss),
    x = "Δt (seconds)",
    y = "Posterior density"
  ) +
  theme_talk()
```

::: {style="font-size: 1.05em; text-align:center; margin-top:0.5em;"}
**0.01 s is well inside the bulk** of the distribution.
:::

---

## What the plot says — and doesn't say

::: {.columns}
::: {.column width="50%"}
::: {style="background:#eaf2f8; padding: 1em; border-left: 3px solid #3a7ca5;"}
**What it says**

A margin of 0.01 s is **not anomalous** under
physically plausible drag assumptions.

The folk theory is not absurd.
It is *consistent* with the physics.
:::
:::

::: {.column width="50%"}
::: {style="background:#fdf4e3; padding: 1em; border-left: 3px solid #d9a441;"}
**What it does NOT say**

The beard *caused* Mieto to lose.

We have no counterfactual race
without the beard.
We have a *distribution* of
plausible drag-time costs.
:::
:::
:::

::: {.fragment style="margin-top: 1.5em; text-align:center; font-size:1.05em; color:#0b1d2e;"}
The right Bayesian sentence: *"0.01 s is consistent with this model."*
The wrong one: *"the beard cost him gold."*
:::

---

## The general lesson

This is a **template** for a very common task in applied data science:

::: {.incremental}
- You have a small effect.
- Someone tells you it is "negligible."
- You have no clean experiment available.
- You have **physics, plausible ranges, and R**.
:::

::: {.fragment style="margin-top:1.5em;"}
Forward simulation lets you ask:

> *Is the observed effect within the distribution of outcomes that
> the system can plausibly produce?*
:::

::: {.fragment style="margin-top:1em; color:#3a7ca5;"}
You don't need a posterior over parameters.
You need a posterior over **consequences**.
:::

---

## Where this approach earns its keep

::: {style="font-size: 0.95em;"}
| Domain | "Negligible" effect | Forward simulation question |
|:-------|:--------------------|:-----------------------------|
| Finance | A 2 bp execution slippage | What return distribution would erase it? |
| Compliance | A rule threshold tweak | How many flagged cases shift, plausibly? |
| Manufacturing | A 0.1 °C drift | Yield consequence under realistic load? |
| Healthcare | A 1 % adherence change | Population-level outcome distribution? |
| Sports | 0.01 s | Drag, wax, draft, terrain — which matters? |
:::

::: {.fragment style="margin-top:1em; text-align:center; color:#0b1d2e;"}
Every row is the **same R script** with different priors.
:::

---

## Why R, specifically

::: {.columns}
::: {.column width="50%"}
**The model fits on one screen.**

- `runif`, `rnorm`, a `for` loop, `ggplot2`.
- No probabilistic programming framework needed for the forward pass.
- Reviewers can read it in 30 seconds.
:::

::: {.column width="50%"}
**The plot is the argument.**

- `geom_density` + `geom_vline` =
the entire claim.
- Reproducible from one `.qmd` file.
- Priors visible, code visible, conclusion visible.
:::
:::

::: {.fragment style="margin-top:1.5em; text-align:center; font-size:1.1em; color:#3a7ca5;"}
**Transparency is the deliverable.** The number is secondary.
:::

---

## Honest limitations

::: {.incremental}
- **Course profile is a proxy.** The 1980 Mt. Van Hoevenberg trail is approximated by today's terrain mix (40/30/30).
- **Weather is assumed calm.** Wind would dominate any beard effect.
- **The `alpha` slipstream factor** is a back-of-envelope shielding estimate, not a CFD result.
- **`p_active`** — the fraction of the race in beard-exposed posture — is the weakest prior.
- **No correlation structure** between the four uncertain parameters.
:::

::: {.fragment style="margin-top: 1em;"}
Every one of these is a knob. Every knob is a `runif(...)` away from being tested.
:::

---

## Extensions (homework, if you like)

::: {.incremental}
- **Sensitivity analysis**: which prior moves $P(\Delta t \geq 0.01\,s)$ most?
- **Replace `runif` with empirical distributions** for $C_d$ from wind-tunnel literature.
- **Add wind.** A 1 m/s headwind dwarfs the beard.
- **Add Wassberg.** Two skiers, two posteriors, one difference distribution.
- **Hierarchical version**: pool across the four cross-country events at Lake Placid.
:::

::: {.fragment style="margin-top:1em; color:#3a7ca5;"}
All of this is plain R. None of it requires more than 100 lines.
:::

---

## Takeaways

::: {style="font-size: 1.05em;"}

1. **Forward simulation is underused.**
   When data is sparse but physics is known, simulate first, fit later.

2. **A prior is a hypothesis.**
   Wide priors are not weakness — they are an honest test of robustness.

3. **The plot is the argument.**
   A single density with a single dashed line settled a 46-year-old question.

:::

::: {.fragment style="margin-top:1.5em; text-align:center; font-size:1.2em; color:#0b1d2e;"}
*A negligible margin is only negligible relative to a model.*
:::

---

## {background-color="#0b1d2e"}

::: {style="color:#c9d6df; text-align:center; padding-top:1em;"}

<h2 style="color:#ffffff;">Thank you.</h2>

<p style="font-size:1.15em; margin-top:1em;">
Kristian Vepsäläinen
</p>

<p style="opacity:0.85;">
M.Sc. Mathematics · M.Eng. Cyber Security<br>
Data scientist · open-data & simulation specialist
</p>

<br>

<p style="font-size:1.05em;">
🌐 <a href="https://www.kristianvepsalainen.com" style="color:#d9a441;">kristianvepsalainen.com</a><br>
✉️ kristian.vepsalainen@proton.me
</p>

<br>

<p style="opacity:0.7; font-size:0.9em;">
Code & slides: github.com/kristianvepsalainen/useR2026 &nbsp;·&nbsp; CC BY 4.0
</p>

:::
